Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest

### 2022-08-05 article/abs.Rmd
draft disc
full draft
rev meth,intro MES need disc,abs
rev intro for res

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
This study addressed how wet tropical secondary forest regeneration was affected by distance to habitat edge, and hypothesized that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.

<!-- # Results/Conclusions -->
As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ca 10% variation explained, while biomass and canopy light tended to stay the same.
Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably. 
Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

@@

# Abstracto{-}

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa arboreal y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.

>>>>

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests. 
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely. 
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration.
However, most studies assess how degraded edges affect intact forests, leaving it unclear whether higher-quality edges could facilitate regeneration of previously degraded forests.
Notably, higher quality edges near intact forests could promote processes like dispersal and wood biomass accumulation that effectively accelerate succession, leading to better performance of shade-tolerant taxa compared to pioneer taxa in the early stages of forested plantation regeneration.
This study addressed how wet tropical forested plantation regeneration was affected by distance to adjacent intact forest edge.
It was hypothesized that old timber plantations facilitate regeneration by increasing available shade, favoring the presence and biomass of later-successional taxa, ultimately changing community composition overall.
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to broad dispersal mode and wood density traits using relevant literature.

<!-- # Results/Conclusions -->
As distance from primary forest edge increased, stem and wood density tended to increase significantly, with ~10% variation explained, while biomass and canopy light tended to stay the same.
Stand tree richness also tended to increase significantly, but diversity decreased steeply and non-linearly, and taxonomic composition varied notably. 
Finally, tree taxa associated with both early and late successional stages decreased significantly, but biomass by dispersal mode did not tend to change.
Overall this study supports that stand composition is less resilient and more subject to edge effects than biomass and richness, suggesting that global forests will likely be distinctly new assemblages in the future, with timber and biodiversity trade-offs occurring based on local and regional management activity.

@@

# Resumen*{-}

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.

### 2022-08-05 article/intro.Rmd
rev intro for res

Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*.
While tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and functionally degraded *[@hubau20]*.
This likely amplifies amid increasing climate stressors *[@anderegg22]* and collective management issues, including insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass >10x faster than old-growth forests *[@finegan96; @poorter16]*, and substantially offset carbon emissions, with estimates ranging from ~10% from Amazonian deforestation *[@smith20]* to a decade of fossil fuel emissions from all of Latin American and the Caribbean *[@chazdon16]*.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80% of old-growth forest levels in just 20 years, especially in wetter regions *[@rozendaal19]*.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
However, recovery of secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]* with added variability *[@atkinson22]*.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, 
**yet local management remains the strongest driver of tropical forest carbon stores *[@pyles22]*.**

Secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
Edge effects also have the potential to shape community composition, based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.

Properties of the existing tree community also depend on land-use history and previous management *[@omeja12]*, together with the composition of surrounding forests.
Overall, legacy effects from management like intensive timber planting could potentially delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
More specifically, timber plantations have historically favored tree taxa that tend to be pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
This could lead to a longer-term dominance of pioneer taxa, referred to as arrested succession *[@]*.
However, arrested succession could be prevented and even typical succession accelerated if intact forests that were adjacent to regenerating plantations functioned as contributing reservoirs of seeds from later-successional tree taxa.
Additionally, the leftover timber stands could provide additional shade, which would not be available in a clear-cut harvested timber plantation, and this could help more shade-tolerant taxa establish and recruit better.
In this case, having a habitat edge near a relatively rich intact forest, rather than near a relatively degraded open area, would instead be an asset to regenerating plantation forest patches, rather than a detriment.
Overall, edge effects near degraded habitat edges are often negative, but influences from meta-population *[@levins79]* and meta-community *[@leibold04; @warren15]* theories do highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity, thereby potentially making particular edge effects, specifically from adjacent intact forests, actually beneficial for regeneration plantation forest parcels.
Taken together, these ideas suggest that near primary forest, shade-tolerant dispersal may facilitate plantation forest regeneration.
This kind of edge effect may ultimately help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.

@@

This study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, highlighting edge effects from an adjacent primary forest.
**- what is new in this study?**
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade, which would suppress less shade-tolerant pioneer tree taxa, and in turn facilitate more shade-tolerant tree taxa, ultimately allowing for more tree diversity within the regenerating plantation forest in parts that were closer to the richer edge of the intact unmanaged forest.
    <!-- - supported by inference?! -->
Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

>>>>

Forest and landscape restoration is a key international conservation and climate change adaptation strategy *[@depinto20]*.
While tropical forests specifically store most land biomass *[@pan13]*, most are now secondary *[@fao20]* and functionally degraded *[@hubau20]*.
This likely amplifies amid increasing climate stressors *[@anderegg22]* and collective management issues, including insufficient policy support *[@chazdon18]* via skewed priorities *[@pyron22]*.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass >10x faster than old-growth forests *[@finegan96; @poorter16]*, and substantially offset carbon emissions, with estimates ranging from ~10% from Amazonian deforestation *[@smith20]* to a decade of fossil fuel emissions from all of Latin American and the Caribbean *[@chazdon16]*.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80% of old-growth forest levels in just 20 years, especially in wetter regions *[@rozendaal19]*.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services *[@guariguata01; @stanturf14]* like biomass and carbon storage *[@liu18]*.
However, recovery of secondary neotropical forests diverge widely in taxonomic composition *[@norden15]*, potentially taking over a century to recover *[@poorter21]* with added variability *[@atkinson22]*.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa *[@carlo16]*, which can be keystone to locally-adapted food webs, including birds *[@maas16]*.
Overall, understanding secondary forest regeneration is key for global biodiversity conservation at the global scale.

Modern secondary forest regeneration is often affected by edge effects, or changes near an ecosystem boundary, especially in heterogeneously-managed landscapes *[@perfecto09; @melo13]*.
About 70% of all forests have been estimated to be within a km of their edge *[@haddad15]*, resulting in tropical forests having been estimated to lose ~22% aboveground carbon along heavily-managed edges >100m, in part due to shortening trees and lighter leaf mass *[@ordway20]*.
Effects of degraded edges have historically favored tree taxa that tend to be pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant *[@tabarelli08]*.
Accordingly, secondary forest wood density may also tend to decrease near degraded habitat edges.
Additionally, the dominance of pioneering taxa could lead to arrested succession *[@tymen15]*, which further delays the recruitment of shade-tolerant taxa, as well as associated increases in aboveground biomass, in part due to their higher survival with increasing shade, and tendency toward relatively higher wood densities.
However, arrested succession could be prevented, and even typical succession accelerated, if intact forests that were adjacent to regenerating plantations functioned as contributing reservoirs of seeds from later-successional tree taxa.
In this case, having a habitat edge near a relatively rich intact forest, rather than near a relatively degraded open area, would instead be an asset to regenerating plantation forest patches, rather than a detriment.
Overall, edge effects ultimately have the potential to shape community composition, in part based on the local dispersal kernel patterns of existing taxa *[@wandrag17]*, which while complicated *[@muller-landau09]*, can also correlate with simpler plant traits like height and seed mass *[@thomson11; @tamme14]*.

Additionally, properties of the tropical forest carbon stores and tree community depend heavily on land-use history and previous management *[@omeja12; @pyles22]*.
Overall, legacy effects from management like intensive timber planting could potentially delay or result in non-linear changes between successional stages *[@albrich20; @gough21]*, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry *[@lefland18]*.
In some cases, the leftover timber stands could provide additional shade, which would not be available in a clear-cut harvested timber plantation, and this could help more shade-tolerant taxa establish and recruit better.
Overall, edge effects near degraded habitat edges are often studied as negative, but conceptual influences from meta-population *[@levins79]* and meta-community *[@leibold04; @warren15]* theories do highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity, thereby potentially making particular edge effects, specifically from adjacent intact forests, actually beneficial for regeneration plantation forest parcels.
Taken together, these ideas suggest that edge effects specifically from adjacent intact primary forests, potentially via shade-tolerant dispersal and performance, may indeed facilitate plantation forest regeneration.
Such a beneficial edge effect may ultimately help explain relatively high biodiversity, compared to adjacent stands, found in agroforests *[@oliveira-neto17]*, forests affected by logging *[@clark12; @edwards14]*, and timber plantations *[@pryde15]*.

@@

Influenced by restoration ecology for forest management, this study reports tree regeneration patterns of an abandoned wet neotropical timber plantation, focused on highlighting edge effects from an adjacent primary forest on overall stand properties, tree community composition, and limited functional trait recovery.
<!-- **- what is new in this study?** -->
We hypothesized that remaining timber trees would mediate forest regrowth by maintaining shade, which would suppress less shade-tolerant pioneer tree taxa, and in turn facilitate more shade-tolerant tree taxa, ultimately allowing for more tree diversity within the regenerating plantation forest in parts that were closer to the richer edge of the intact unmanaged forest.
    <!-- - supported by inference?! -->
Accordingly, we predicted that further from the primary forest edge, canopy light availability would increase, while stand biomass and diversity would decrease.

### 2022-08-05 article/merge.Rmd
rev intro for res

<!-- journals: J [Appl] Ecol (6.8), J [Appl] Veg Sci (3.4), Basic Appl Ecol (3.7), Biotropica (2.5), J Forestry Res (2.2), Small-scale Forestry (1.3), Eco Evo -->

>>>>

<!-- journals: J [Appl] Ecol (6.8), J [Appl] Veg Sci (3.4), Basic Appl Ecol (3.7), Biotropica (2.5), J Forestry Res (2.2), J Trop Ecol (1.3), Small-scale Forestry (1.3), Eco Evo -->
